Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-17 forecastflows.tex
that's all for today

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. % jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
If a model becomes too complex, it may become a black box, and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}. % jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty .... [FLAG: agree but not really relevant to PBM? Maybe it's not clear enough we are focusing on PBM in this section...]

Applying the workflow to process-based models would help open the black box. % jdavies15april: what is the black box?
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and help expose potential degeneracies in the model design. %jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below)
Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight degeneracies by requiring more explicit model calibration, % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated). Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections. 
% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties? This could help illustrate potential degeneracies etc.

@@

% Trends - outline current problem
%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology. 
% Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \citep{soule1999conserving,soule1991conservation}. The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}. 
% Such trends drove a number of new subfields within ecology---some of which are now complete disciplines within themselves \citep[such as conservation biology,][]{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES. Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity. 
% Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. 
In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results. % jdavies16april: could highlight how influential the LPI has become (recent reports suggest a 73% ave. pop decline since 1970. This would be massive!
With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}. 
% jdavies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations. We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends. For this we need to predict outside of the set ...

% Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines, and has led to a debates within ecology about whether such analyses undermine public confidence in science  \citep{gonzalez2016estimating}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.

% Functionally, research on the LPI has somewhat reverse-engineered the recommended workflow: after a series of papers debating different estimates from different models, more recent papers have focused on simulated data to highlight uncertainty given the model and data togethers \citep[though I don't think they link their simulations to the model they use that well,][]{dove2023quantifying,toszogyova2024mathematical}, but this should have been part of the process for the very first papers.
% While this debate captures part of the process of science---refining our approaches and thus estimates over time---it also makes it hard for policy-makers to do (?? something) and has led to a debates within ecology about whether such analyses undermine public confidence in science  (CITES Biodiversity debate). 
% \citet{Buschke2021} who focused on adding random population variation
\vfill

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another. % jdavies15april: is this more specific to process based  models?

@@

\centerline{\bf A common workflow to bridge trend estimation and forecasting} % jdavies15april: would it be possible to label this panel with components of the LPI to provide a clearer illustration of how the framework is implemented? % vvdm16april: not sure I understood
\vspace*{-3mm}
{\begin{multicols}{2}
\begin{minipage}[t]{\linewidth}
\includegraphics[width=\linewidth]{figures/iterativeworkflow_details}

@@

We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed. This may be especially true for those who see science rewarded mostly through the shear number of publications. But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help. % jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).

>>>>

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. % jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
If a model becomes too complex, it may become a black box, % mention black box here to line up with the beginning of next paragraph
and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}. % jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty .... [FLAG: agree but not really relevant to PBM? Maybe it's not clear enough we are focusing on PBM in this section...]

Applying the workflow to process-based models would help open the black box. % jdavies15april: what is the black box?
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential degeneracies in the model design. %jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below) [FLAG: agree, and made me realize it was not 100% clear in my mind. As I understand it, non-identifiability concerns one parameter, and degeneracy concerns the all model?... but not sure]
Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight degeneracies by requiring more explicit model calibration, % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated). Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections. 
% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties? This could help illustrate potential degeneracies etc.

@@

% Trends - outline current problem

In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results. % jdavies16april: could highlight how influential the LPI has become (recent reports suggest a 73% ave. pop decline since 1970. This would be massive!
With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}. 
% jdavies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations. We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends. For this we need to predict outside of the set ... [FLAG]

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another. % jdavies15april: is this more specific to process based  models? [FLAG: hmm yes again? Same point as above, maybe it's not clear we're talking mostly about PBM?]

@@

\centerline{\bf A common workflow to bridge trend estimation and forecasting} % jdavies15april: would it be possible to label this panel with components of the LPI to provide a clearer illustration of how the framework is implemented? % [FLAG: not sure I understood]
\vspace*{-3mm}
{\begin{multicols}{2}
\begin{minipage}[t]{\linewidth}
\includegraphics[width=\linewidth]{figures/iterativeworkflow_details}

@@

We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
% One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed. This may be especially true for those who see science rewarded mostly through the shear number of publications. But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help. % jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
Good model fitting is inherently iterative and takes time, at odds with the pressure to publish quickly.
Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).
